---
title: "Analyzing Spatial Data with R - exercises"
author: "Micha Silver"
date: "2025-06-03"
output:
  html_document: default
  pdf_document: default
bibliography: references.bib  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r pkgs, results='hide', message=FALSE}
pkg_list <- c("terra", "sf", "CDSE", "knitr", "leaflet", "ggplot2")
invisible(lapply(pkg_list, library, character.only = TRUE))
```
## Get CDSE credentials

and prepare OAuth token for API access

```{r creds, results='hide'}
creds <- read.csv("credentials.csv")
tok <- CDSE::GetOAuthToken(id = creds$clientid, secret = creds$secret)
```
### Set query parameters

- Dates
- AOI file
- Which "collection" (Sentinel-2 Level L2A)

```{r params}
aoi <- sf::st_read(file.path("GIS", "Fuenteduque_aoi.gpkg"))
from_date <- "2025-01-01" 
to_date <- "2025-04-30"
collection <- "sentinel-2-l2a"
max_cloud <- 20
tileid <- "29SQB"  # Download only one Copernicus tile
```
### Get list of available images

```{r catalog}
img_list <- CDSE::SearchCatalog(aoi = aoi,
                                from = from_date,
                                to = to_date,
                                collection = collection,
                                token = tok)
# How many images available?
message("Number of available images: ", nrow(img_list))
```
### Remove images with high cloud cover

#### Also select only one image tile, "29SQB"

```{r cloud-filter}
img_list <- img_list[img_list$tileCloudCover <= max_cloud,]
img_list <- img_list[grepl(tileid, img_list$sourceId),]
message("Number of images after cloud filtering: ", nrow(img_list))
# Sort by dates
img_list <- img_list[order(img_list$acquisitionDate),]
# Which image dates are available?
knitr::kable(img_list$acquisitionDate)
```
### Loop over available images

```{r water-surface}
# Setup Output directory
Output_dir <- "Output"
if (!dir.exists(Output_dir)) {
  dir.create(Output_dir)
}

# We already have collection from above;
# Refresh token:
tok <- CDSE::GetOAuthToken(id = creds$clientid, secret = creds$secret)

# Use lapply to loop over img_list data.frame:
water_rast_list <- lapply(1:nrow(img_list), function(i) {
  dte <- img_list$acquisitionDate[i]
  
  # Save processed water surface as Geotiff
  rast_file <- file.path(Output_dir, paste0("water_", dte, ".tif"))
  # But, if file already exists (from previous run), just skip download
  if (!file.exists(rast_file)) {
  water_rast <- CDSE::GetImage(aoi = aoi,
                               time_range = dte,
                               collection = collection,
                               script = "MNDWI_masked.js",
                               resolution = 20, # Use lower res of SWIR band
                               token = tok)
  
  terra::writeRaster(water_rast, rast_file)
  } else {
    #message("Raster file: ", rast_file, " already exists, skipping.")
  }
  return(rast_file)
})
water_rast_list <- unlist(water_rast_list)
message("Saved: ", length(water_rast_list), " water surface raster files.")
```
### Plot water surface map

```{r plot1}
# Take first water surface map in list
r <- rast(water_rast_list[1])
leaflet() |> addTiles() |>
  addRasterImage(r)
```

### Threshold to get just water surfaces

#### Using the threshold value 0.09 (@xu_modification_2006)

```{r water-thresh}
thresh <- 0.09
reclass_mat <- matrix(c(-1.0, thresh, NA, thresh, 1.0, 1), ncol=3, byrow=TRUE)
reclass_mat
water_thresh_list <- lapply(water_rast_list, function(x) {
  r <- terra::classify(rast(x), reclass_mat)
  water_thresh_file <- gsub("water", "water_thresh", x)
  terra::writeRaster(r, water_thresh_file, overwrite = TRUE)
  return(water_thresh_file)
})
water_thresh_list <- unlist(water_thresh_list)
```

### Plot thresholded water surface

```{r water-thresh1}
# Take first water surface map in list
thresh_file <- water_thresh_list[1]
r <- rast(thresh_file)
leaflet() |> 
  addTiles() |>
  addRasterImage(r)
```

```{r water-thresh2}
# and the second image
thresh_file <- water_thresh_list[2]
r <- rast(thresh_file)
leaflet() |> 
  addTiles() |>
  addRasterImage(r)
```


```{r water-thresh3}
# Now the last water surface map in list
thresh_file <- water_thresh_list[length(water_thresh_list)]
r <- rast(thresh_file)
leaflet() |>
  addTiles() |>
  addRasterImage(r)
```

### Extract value at point for all water rasters

#### Choose a sampling point

```{r sample-point}
# Show the point on map
coords <- data.frame("lon" = -6.3710,
                     "lat" = 37.0375)
pt <- vect(coords, crs = "EPSG:4326")
r <- rast(water_rast_list[2])
leaflet() |>
  addTiles() |>
  addRasterImage(r) |>
  addCircleMarkers(lng = coords$lon, lat = coords$lat, radius = 7,
                   fillColor = "red", stroke = FALSE, fillOpacity = 1)
```

#### Extract time series of MNDWI values at that point

```{r extract}
water_val_list <- lapply(water_rast_list, function(x) {
  r <- rast(x)
  val <- terra::extract(r, pt, ID = FALSE)
  # Remove extra file name character, leave only date for plotting
  dte <- basename(x)
  dte <- gsub("water_", "", dte)
  dte <- gsub(".tif", "", dte)
  val_df <- data.frame(as.Date(dte), val)
  names(val_df) <- c("Image_date", "MNDWI")
  return(val_df)
})
# Bind list into data.frame
vals_df <- do.call(rbind, water_val_list)
knitr::kable(vals_df)
```

#### Time series plit

```{r plot-timeseries}
ggplot(vals_df) +
  geom_col(aes(x=Image_date, y=MNDWI), fill = "blue") +
  geom_hline(aes(yintercept = 0.09,
            color = "red") ) +
  ggtitle("MNDWI values",
          subtitle = paste("at location", coords$lon, coords$lat)) +
  theme(axis.text.x = element_text(angle = 30),
        legend.position = "")
```

## License

eLTER Workshop-Analyzing Remote Sensing Data in R 

© 2025 by Micha Silver

is licensed under Creative Commons Attribution-ShareAlike 4.0 International.

To view a copy of this license, visit

https://creativecommons.org/licenses/by-sa/4.0/

## References